Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-06-14 SIMPaper/SIMPaper.Rmd
added acknowledgement

This work is supported through 5-UM1-DK078616‐14 and R01-HL151855-01 grants.

>>>>

This work was supported by subcontracts from UM1DK078616 and R01HL151855 to RS. We would like to thank Dr. James Meigs, Project Leader of these awards, for his support and helpful discussions. We would also like to Thank Dr. James Hanley for his support and discussions while extending the case-base methodology.

### 2022-06-14 SIMPaper/SIMPaper.tex
added acknowledgement

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While these methods have led
to better predictive performance than regression-based approaches, they
cannot model both time-varying interactions and complex baseline
hazards. To address this, we propose Case-Base Neural Networks (CBNN) as
a new approach that combines the case-base sampling framework with
flexible architectures. Our method naturally accounts for censoring,
does not require additional hyperparameters and uses binary
cross-entropy as the loss function. Using a novel sampling scheme and
data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. This study highlights the benefit of
combining case-base sampling with deep learning to provide a simple and
flexible modeling framework for data-driven, time-varying interaction
modeling of survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2\)
and \(\psi\) are basis splines. To generate the \(gamma\) coefficients,
an initial cubic splines model with three knots is fit using the
\emph{flexsurvspline} function with an intercept-only model on the data,
which returns the coefficients (\(\gamma\)) we use in our model. Note
that we fix these values for the analysis. The \(\beta\) coefficients
represent direct effects, \(\tau_{2}\) and \({\tau_3}\) represent
interactions and \(\tau_{1}\) is a time-varying interaction.

@@

This work is supported through 5-UM1-DK078616‐14 and R01-HL151855-01
grants.

>>>>

\abstract{Neural network-based survival methods have been developed to
model data-driven covariate interactions. While these methods have led
to better predictive performance than regression-based approaches, they
cannot model both time-varying interactions and complex baseline
hazards. To address this, we propose Case-Base Neural Networks (CBNN) as
a new approach that combines the case-base sampling framework with
flexible architectures. Our method naturally accounts for censoring,
does not require additional hyperparameters and uses binary
cross-entropy as the loss function. Using a novel sampling scheme and
data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in two simulations and two real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a simple simulation with an exponential hazard model,
CBNN outperforms the other neural network methods. For a complex
simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions, CBNN outperforms all
competitors. The first real data application shows CBNN outperforming
all neural network competitors, while a second real data application
shows competitive performance. We highlight the benefit of combining
case-base sampling with deep learning to provide a simple and flexible
modeling framework for data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

where
\(\gamma_{1}=3.9, \gamma_{2}=3, \gamma_{3}=-0.43, \gamma_{4}=1.33,\gamma_{5}=-0.86, \beta_{{1}}=1, \beta_{{2}}=1, \beta_{{3}}=1, \tau_{1}=10, \tau_{2}=2, \tau_{3}=2\)
and \(\psi\) are basis splines. The \(gamma\) coefficients are obtained
from an intercept-only cubic splines model with three knots using the
\emph{flexsurvspline} function from the \textbf{flexsurv} package
\citep{flexsurv}. Note that we fix these values for the analysis. The
\(\beta\) coefficients represent direct effects, \(\tau_{2}\) and
\({\tau_3}\) represent interactions and \(\tau_{1}\) is a time-varying
interaction.

@@

This work was supported by subcontracts from UM1DK078616 and R01HL151855
to RS. We would like to thank Dr.~James Meigs, Project Leader of these
awards, for his support and helpful discussions. We would also like to
Thank Dr.~James Hanley for his support and discussions while extending
the case-base methodology.
